---
title: "lab_02"
author: "derek willis"
date: "8/24/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## About this lab

To complete this lab, you need to:
* run existing code as directed (look for **Task**).
* modify existing code as directed (look for **Task**).
* write code in empty codeblocks provided to answer questions included (look for **Q**).
* write out the answer in the form of a complete sentence in the space given (look for **A**).

When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

## Load libraries and establish settings

**Task** Run the codeblock below to load the Tidyverse and Lubridate. To run the codeblock, hit the little green play button in the upper right corner of the gray area (the codeblock) below.

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse and lubridate
library(tidyverse)
library(lubridate)
```

## Load Data

Let's keep working with West Virginia loans contained in the PPP loans data.  

**Task** Run the codeblock below to load the data.

```{r}
west_virginia_ppp <- read_rds("data/lab_02.rds")
```

## Examine Data

## Answer questions

**Q1** Create a new column that divides the loan amounts into the following options - under_10k, 10k_50k, 50k_100k, 100k_500k, 500k_1m, 1m_plus - and calculates how many loans are in each group. How would you describe the distribution of loans in a sentence?
Hint: You will need to use `case_when` to make this work.
**A1** Most of the loans were under /$50,000; less than 1,000 of the states 33,000+ loans were for over /$500,000.

```{r}
west_virginia_loan_amount <- west_virginia_ppp %>% mutate(loan_amount_category = case_when(
  amount < 10000 ~ 'under_10k',
  amount >= 10000 & amount < 50000 ~ '10k_50k',
  amount >= 50000 & amount < 100000 ~ '50k_100k',
  amount >= 100000 & amount < 500000 ~ '100k_500k',
  amount >= 500000 & amount < 1000000 ~ '500k_1m',
  amount >= 1000000 ~ '1m_plus'
)) 

west_virginia_loan_amount %>% group_by(loan_amount_category) %>% summarize(count=n()) %>% arrange (loan_amount_category)

```

**Q2** Some of the loans appear not to have been paid out - there is a column representing the amount of money that hasn't been disbursed to the applicant. Which servicing lenders have the most loans with undisbursed amounts greater than 0?
**A2** Davis Trust Company holds twelve of the thirteen loans that still have outstanding funds that need to be paid to recipients.

```{r}
west_virginia_undisbursed <- west_virginia_ppp %>% filter(undisbursed_amount > 0)

```

**Q3** What are some potential reasons why a loan might not be paid out? What questions does that circumstance raise for you?
**A3** I can think of several reasons. For one, the beneficiary may have gone out of business. Maybe the loans were found to be fraudulent - but it seems improbable that this would be caught before the loan was paid out. It is also possible that the issuing lender might have gone under, or be experiencing some sort of financial insolvency. This seems to me to be the likeliest scenario. In this case, I would want to know what caused this to happen. Did the PPP loans put the lender too far into the red? Is the government putting too much pressure on smaller financial institutions to help them enact their mandate?


**Q4** Which month in 2021 had the most loans approved?
**A4** March had the most loans approved in 2021.

```{r}
# Write the code to produce your answer here

west_virginia_months <- west_virginia_ppp %>% 
  mutate (CleanDate = ymd(date_approved)) %>% 
  mutate (Month = floor_date(CleanDate, "month")) %>%
  group_by(Month) %>%
  summarize(total=n()) %>%
  arrange(desc(total))

```

**Q5** How many loans went to West Virginia applicants that listed the maximum number of jobs retained (500)? Which one had the earliest approved date?
**A5** 10 companies listed that they retained the maximum of 500 jobs. J. Austin Heating & Cooling Inc. was approved first, on April 8, 2020.

```{r}
west_virginia_ppp %>% 
  filter(jobs_retained == 500) %>%
  arrange (date_approved)

```
