---
title: "lab_03"
author: "sean mussenden"
date: "8/19/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings
```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse here
library(tidyverse)
# Load janitor here
library(janitor)
```

## Load data
```{r}
# Write code to load the WV slice of the ppp loan data here and call it wv_ppp_loans.
# The data is called lab_03.rds, and is in the same folder as lab_03.Rmd (this data notebook)
read_rds("lab_03.rds") -> wv_ppp_loans

```

## Answer questions

Q1: Which five cities located in Kanawha County, West Virginia, got the largest _number_ of loans? How many did each get? To answer this question, you'll have to clean the very messy "city" field.

Note: To answer this question, you'll need to see every variation of the "city" column in Kanawha County, and then fully clean the city field for those cities. Hint: once cleaned, the dataframe that counts the number of loans by city should have 38 distinct cities for Kanawha County.

A1: Charleston (2802), South Charleston (478), Saint Albans (296), Dunbar (162), Nitro (140)
```{r}

kanawha_ppp_loans <- wv_ppp_loans %>%
  filter(project_county_name == "KANAWHA") %>% 
  mutate(city = str_to_title(city)) %>% 
  mutate(city = case_when(
      str_detect(city, "^St") ~ "Saint Albans",
      str_detect(city, "^So") ~ "South Charleston",
      str_detect(city, "^Elk") ~ "Elkview",
      str_detect(city, "^Cros") ~ "Crosslanes",
      city == "Balle" ~ "Belle",
      city == "S Charleston" ~ "South Charleston",
      city == "Charelston" ~ "Charleston",
      TRUE ~ city
)) %>% 
  group_by(city) %>% 
  summarise(count_loans=n()) %>%
  arrange(desc(count_loans))

```

Q2: Find the 5-digit ZIP code in West Virginia that had the highest total amount of loans (by dollar value). What is the name of the business that got the largest loan (by dollar value) in that zip code? Write up a short paragraph that also answers these questions:

* Where in West Virginia is this ZIP code located?
* How much was the business' loan for?
* How much of that loan was forgiven?
* Which number is bigger, the loan amount or the forgiveness amount? Can you explain the discrepancy?

A2:
25301 is in Charleston. The business, National Travel Service, Inc, was loaned 749150, and forgiven 754350.35. I think this is due to inflation accrued by the loan.
```{r}

wv_zip_loans <- wv_ppp_loans %>% 
  mutate(zip = str_sub(zip, start=1L, end=5L)) %>% 
  group_by(zip) %>% 
  summarise(total = sum(amount)) %>% 
  arrange(desc(total))

wv_ppp_loans %>% 
  mutate(zip = str_sub(zip, start=1L, end=5L)) %>%
  filter(zip == "25301") %>% 
  arrange (desc(amount))
```

Q3: Five business entities owned by the spouse of a member of West Virginia's U.S. House delegation got PPP loans. Write up a paragraph that answers the following questions.   

  * What are the businesses' names?
  * What do the businesses do?
  * How much money did the five businesses receive in total?
  * What percentage of the total loan amount have been forgiven?
  * How did the U.S. House member vote on the legislation initially author zing the PPP loan program.

Feel free to add any interesting "color" (little interesting details) you find in your research.  

Some hints to help you answer this question:

* West Virginia's three U.S. House members are listed [here](https://www.govtrack.us/congress/members/WV#representatives)
* To find businesses they or their families own, you can search for financial disclosure forms for U.S. House members [here](https://disclosures-clerk.house.gov/PublicDisclosure/FinancialDisclosure#Search).
  * To speed things up, here are direct links to the three U.S. House members most recent financial disclosures, which I found by searching the site linked above.
  * [Rep. 1](https://disclosures-clerk.house.gov/public_disc/financial-pdfs/2020/10042042.pdf)
  * [Rep. 2](https://disclosures-clerk.house.gov/public_disc/financial-pdfs/2020/10040330.pdf)
  * [Rep. 3](https://disclosures-clerk.house.gov/public_disc/financial-pdfs/2020/10042734.pdf)
* When you examine the disclosure forms, there may be lots of stock sales and retirement accounts mixed in.  Ignore those. In the table in the disclosure forms, look for "Partnership" or "C-Corp" in the column named "Income Type"
* This is a link to details on passage of the [CARES ACT](https://www.congress.gov/bill/116th-congress/house-bill/748). You can find a link to how the House voted by clicking on a link from [this page](https://www.congress.gov/bill/116th-congress/house-bill/748/actions).

A3:
US Rep Carol Devine Miller (R-WV3) has a husband, Dutch, who is in the car business. He owns five dealerships: four in Hungtington, one in Charleston. All together, Dutch's businesses received 3105300 dollars, of which 3140471 dollars were forgiven. That's 100% plus interest. Rep. Miller, interesting enough, voted "YEA" on the passage of the CARES Act. More interestingly, Dutch is a veteran, yet none of his businesses are classified as veteran-owned on the PPP datasheet. Most interestingly, there are way too many adult males named Dutch in or adjacent to US Politics.
```{r}

# Hint
# First, create a dataframe that only has loans to those five companies
# Second, use that newly created dataframe to calculate total loan amount and percentage forgiven for those companies

usrep_ppp_loans <- wv_ppp_loans %>% 
  filter(str_detect(name, "^DUTCH MILLER") | str_detect(name, "^DM MO"))

usrep_ppp_loans %>% 
  summarise(
    sum(amount)
)

usrep_ppp_loans %>% 
  summarise(
    sum(forgiveness_amount)
  )

```
