---
title: "lab_06"
author: "derek willis"
date: "8/26/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## About this lab

To complete this lab, you need to:
* [install Tabula](https://tabula.technology/)
* run existing code as directed (look for **Task**).
* write code as directed (look for **Task**).
* write code in empty codeblocks provided to answer questions included (look for **Q**).
* write out the answer in the form of a complete sentence in the space given (look for **A**).

When you are finished, commit changes in the lab_06 folder and push to your personal GitHub repo, then submit the URL to this document on ELMS.

## Load libraries and establish settings

**Task** Run the codeblock below to load the Tidyverse. To run the codeblock, hit the little green play button in the upper right corner of the gray area (the codeblock) below.

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse.
library(tidyverse)
```

## Get Our PDF

We'll be working with the [Maryland Statistical Handbook](https://planning.maryland.gov/MSDC/Documents/md-statistical-handbook.pdf) from the state Department of Planning. It's not tiny (44MB), and you'll want to download it to a place you'll remember (like your Downloads folder).

**Task** Download the file from the above link and save it to your Downloads folder.

## Setup Tabula

**Task** Start Tabula and load the PDF, then extract a table

Start Tabula, then go to http://127.0.0.1:8080/ in your browser. Click the "Browse" button and find the Statistical Handbook file in your Downloads folder and click "open", and then click the "Import button" in Tabula. This will take a minute or two.

Let's go to page 30 of the PDF, Table 2A, "International Migration for Maryland's Jurisdictions, July 1, 2010 to July 1, 2019". This is _net migration_, meaning it shows the total number of people moving to Maryland and its regions/jurisdictions _minus_ the total number of people leaving the same, divided into 12-month periods from July to July. In Tabula, draw a box around that table's border and click the "Preview & Export Extracted Data" button. It should look pretty clean. Let's export that CSV (it should be called `tabula-md-statistical-handbook.csv` by default) to your lab_06/data folder.

## Cleaning up the data in R

Let's load it into R, and in doing so we'll skip the first two rows and add our own headers that are cleaner. `read_csv` allows us to do this ([and more](https://readr.tidyverse.org/reference/read_delim.html)):

**Task** Write code to read the CSV file into a dataframe called `international_migration`. As in the pre_lab, you may have to skip 1 or more rows and you may want to supply easier to use column names that use underscores.

```{r}
maryland_migration <- read_csv(
  'data/tabula-md-statistical-handbook.csv',
  skip = 2,
  col_names = c('jurisdiction', '2010_2011', '2011_2012', '2012_2013', '2013_2014', '2014_2015', '2015_2016', '2016_2017', '2017_2018', '2018_2019', 'total')
)
```

Add a column for the type of migration ("international") and populate it:

**Task** Write the code to add a `migration_type` column to the data and give it the value of "international"

```{r}
maryland_migration <- maryland_migration %>% 
  mutate (
    migration_type = 'international'
  )
```

## Answer questions

**Q1** Which region and county/city _below_ the state level accounted for the largest percentage of international migration for the entire time period? You'll need to add and populate a column representing percent of total using `mutate`.
**A1** The Suburban Washington region accounted for 62.8% of migration in Maryland. Montgomery County accounts for the largest single county of migrants received, at 36.8%.

```{r}
maryland_migration_percents <- maryland_migration %>% 
  mutate (
    percent = (total/198996)*100
  )

#Regions
maryland_migration_percents %>% 
  filter (
    str_detect(jurisdiction,' Region'),
    jurisdiction != 'Maryland'
  ) %>% 
  arrange (desc(percent)) %>% 
  select (jurisdiction, percent)

#Counties
maryland_migration_percents %>% 
  filter (
    !str_detect(jurisdiction,' Region'),
    jurisdiction != 'Maryland'
  ) %>% 
  arrange (desc(percent)) %>% 
  select (jurisdiction, percent)
```

**Q2** Write a sentence or two that describes the data you produced in A1. Try to convey the most important idea.
**A2** The data shows what percentage of total migrants to Maryland relocated to each of the different jurisdictions in Maryland. 

**Q3** Which region & jurisdiction had the largest percentage change (either negative or positive) for international migration between July 2018 and July 2017? The formula for percentage change is easy to remember: (New-Old)/Old.
**A3** Migrants relocating to the Southern Maryland region decreased by 140.7%. Calvert County, which is in Southern Maryland, saw the largest single-county decrease with a 333.3% reduction.

```{r}
#Regions
maryland_migration %>% 
  select(jurisdiction,`2016_2017`,`2017_2018`) %>% 
  mutate(
    percent_change = ((`2017_2018`-`2016_2017`)/`2017_2018`)*100
  ) %>% 
  filter (
    str_detect(jurisdiction,' Region'),
    jurisdiction != 'Maryland'
  ) %>% 
  arrange(percent_change)
  

#Counties
maryland_migration %>% 
  select(jurisdiction,`2016_2017`,`2017_2018`) %>% 
  mutate(
    percent_change = ((`2017_2018`-`2016_2017`)/`2017_2018`)*100
  ) %>% 
  filter (
    !str_detect(jurisdiction,' Region'),
    jurisdiction != 'Maryland'
  ) %>% 
  arrange(percent_change)
```

**Q4** What's your best guess as to why these changes occurred, and in those places in particular?
**A4** Data from the second year represents the first full year of the Trump presidency, which was characterized by very strict new immigration policies. Southern Maryland, including Calvert County both went for Trump in the 2016 election, and residents there may have been more gung-ho about the new policies, and made a less welcoming environment for migrants. 

Also, it is worth noting that in areas with small numbers to begin with, percent changes look much larger compared to areas that started with higher number. A loss of 40 migrants in Calvert county is significant when they only had 52 to begin with, but the same number in Prince George's County would have made for a change of less than one percent. 


## Back to Tabula

**Task** Extract second table from Tabula

Let's go to page 31 of the PDF, Table 2B, "Domestic Migration for Maryland's Jurisdictions, July 1, 2010 to July 1, 2019". In Tabula, hit the "Clear All Selections" button and then draw a box around that table's border and click the "Preview & Export Extracted Data" button. It should look pretty clean. Let's export that CSV to your lab_06/data folder. (and rename it to `tabula-md-statistical-handbook-domestic.csv`).

## Cleaning up the data in R

Let's load it into R, and in doing so we'll skip the first two rows and add our own headers that are cleaner:

**Task** Write code to read the CSV file into a dataframe called `domestic_migration`. As in the pre_lab, you may have to skip 1 or more rows and you may want to supply easier to use column names that use underscores. Your work on the international file should serve as a guide.

```{r}
maryland_domestic <- read_csv(
  'data/tabula-md-domestic.csv',
  skip = 2,
  col_names = c('jurisdiction', '2010_2011', '2011_2012', '2012_2013', '2013_2014', '2014_2015', '2015_2016', '2016_2017', '2017_2018', '2018_2019', 'total')
)
```

Add a column for the type of migration ("domestic") and populate it:

**Task** Write the code to add a `migration_type` column to the data and populate it with the value of "domestic"

```{r}
maryland_domestic <- maryland_domestic %>% 
  mutate (
    migration_type = 'domestic'
  )
```

## Answer questions
**Q5** Which Maryland individual jurisdiction saw the largest net decrease in domestic migration overall?
**A5** Baltimore City

```{r}
maryland_domestic %>% 
  filter (
    !str_detect(jurisdiction,' Region'),
    jurisdiction != 'Maryland'
  ) %>%
  arrange(total) %>% 
  select(jurisdiction,total)

```

**Q6** How many regions & jurisdictions had net positive migration for July 2017, July 2018 and July 2019 (not just individually but all three together)?
**A6** 14

```{r}
maryland_domestic %>% 
  filter(
    `2016_2017` > 0 &
    `2017_2018` > 0 &
    `2018_2019` > 0
  ) %>% 
  select(jurisdiction,`2016_2017`,`2017_2018`,`2018_2019`)

```

**Q7** How would you describe this data? Is there a county or region that stands out, and why?
**A7** This shows the less and less Americans in the 2010s have been relocating to Maryland. There is, however, one outlier. The Southern Maryland region saw a significant increase in domestic migrants. Coincidentally, this is also the region that saw the biggest decrease in international migrants under the Trump presidency. It's tempting to think that this is because it's a magnet for xenophobes, but there may very well be other factors at play, like a lower cost of living. Still, it's enough to make you stop and think.