---
title: "lab_04"
author: "derek willis"
date: "8/24/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## About this lab

To complete this lab, you need to:
* run existing code as directed (look for **Task**).
* modify existing code as directed (look for **Task**).
* write code in empty codeblocks provided to answer questions included (look for **Q**).
* write out the answer in the form of a complete sentence in the space given (look for **A**).

When you are finished, commit changes and push to your personal GitHub repo, then submit the URL to this document on ELMS.

## Load libraries and establish settings

**Task** Run the codeblock below to load the Tidyverse. To run the codeblock, hit the little green play button in the upper right corner of the gray area (the codeblock) below.

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse.
library(tidyverse)
library(lubridate)
```

## Load Data

Let's keep working with Maryland loan applications contained in the PPP data.

**Task** Read in the lab_04.rds file and assign it to the variable `maryland_ppp`.

```{r}
maryland_ppp <- read_rds("lab_04.rds")
```

To answer the questions below, you'll need to join both the NAICS code data and the ZCTA demographic data from the pre-lab. There's a particular set of steps you'll need to perform to make that work properly, so take your time and check your work.

First, let's load the NAICS code CSV file we have in the data folder:

**Task** Read in the naics_codes.csv file and assign it to the variable `naics_codes`.

```{r}
naics_codes <- read_csv("data/naics_codes.csv")
```

Next join the two and then create a 5-character `zip5` column that we can use to join to the ZCTA data:

**Task** Write code to join the maryland_ppp and naics_codes data, assigning the result to a new variable. Then add a new `zip5` column.

```{r}
maryland_ppp_naics_zip <- maryland_ppp %>% left_join(naics_codes) %>% mutate(zip5 = str_sub(zip,1,5))
```

Let's add the Zip Code Tabulation Area dataset. You'll need to make the `ZCTA5N` column a character column, not a numeric one.

**Task** Read in the Zip Code Tabulation Area dataset in the data folder and change the datatype of the `ZCTA5N` column to character. IMPORTANT: you can see what the column names represent by checking the data documentation, which can be found here: https://geodata.md.gov/imap/rest/services/Demographics/MD_CensusData/FeatureServer/1. You will need to refer to this to answer the questions below.

```{r}
maryland_zcta <- read_csv("data/maryland_zcta.csv") %>% mutate(across(ZCTA5N,as.character))
```

Now we can join the PPP data to the ZCTA data:

**Task** Join the Maryland PPP data (with NAICS already joined) with the ZCTA data, specifying the join column in both dataframes.

```{r}
maryland_ppp_naics_zcta <- maryland_ppp_naics_zip %>% left_join(maryland_zcta, by=c("zip5"="ZCTA5N"))
```

## Answer questions

**Q1** What are the top 5 industries (using the NAICS title) with the most PPP applications in Maryland?
**A1** Taxi Service, Beauty Salons, Full-Service Restaurants, Office of Real Estate Agents and Brokers, "All Other Personal Services"

```{r}
maryland_ppp_naics_zcta %>% 
  group_by(title) %>% 
  summarize(count_loans=n()) %>% 
  arrange (desc(count_loans))
```

**Q2** What are the top industries (using the NAICS title) with the most PPP applications in Maryland where the loan amount is more than $150,000?
**A2** 
Some of the top industries with businesses that received loans over $150,000 are full-service restaurants, doctors' offices, HVAC and plumbing contractors, dentists' offices, and construction agencies that build commercial and institutional facilities.
```{r}
maryland_ppp_naics_zcta %>% 
  filter(amount > 150000) %>% 
  group_by(title) %>% 
  summarize(count_loans=n()) %>% 
  arrange(desc(count_loans))

```

**Q3** Which Maryland zip code with at least 20 percent non-Hispanic Asian population had the largest sum of PPP loans? Also show the number (count) of loans.
**A3** 20850, in Montgomery County, received a total of $47452058, spread over 3186 separate loans. 

```{r}
maryland_ppp_naics_zcta %>% 
  filter(PNHA > 20) %>% 
  group_by(zip5) %>% 
  summarize(
    total_loans=sum(amount),
    count_loans=n(),
  ) %>% 
  arrange(desc(total_loans)) %>% 
  select("zip5","total_loans","count_loans")

```

**Q4** Which Maryland zip code had the largest sum of loans where the reported race of the applicant was "Asian", and what is the percentage of non-Hispanic Asian population in that zip code? Also show the number (count) of loans.
**A4** 20850 again. Its population is 23.5% Asian.

```{r}
maryland_ppp_naics_zcta %>% 
  filter(race == "Asian") %>% 
  group_by(zip5,PNHA) %>% 
  summarize(
    total_loans=sum(amount),
    count_loans=n()
  ) %>% 
  arrange(desc(total_loans))
```

**Q5** What might explain why 20850 has a significant Asian population (23.5%) and only 240 of the 3,186 loans (7.5%) are to Asian applicants?
**A5** 20850 is Rockville, which is where many wealthy DC workers live. I think it's reasonable to assume more of the residents of Rockville and its suburbs commute to work, either to their owned businesses or to give the surplus value of their labor to a boss, rather than own businesses close to their homes.


**Q6** Check to see if any Maryland zip codes that have loans where the PPP data lists the applicant as being Asian BUT are located in a zip code that does not appear to have Asian population according to the ZCTA data. How many loans do they have and what is the total amount? Take a look at these zip codes - what could explain this?
**A6** There are three zip codes that fit this criteria. 20701 (Annapolis Junction - a silly place, let us never go there) has four loans to Asian-owned businesses totalling $75170.50. 21240 (BWI - yikes) has three, totalling $358674. 21031 (Cockeysville - represeeeeent) has 6 loans totalling... close to $7m. That's wild to me... but I guess Pharmaceuticals Int'l needed a lot of cash.

All three of these areas do not have residential space (their combined population is only 6), so there are barely any people there to begin with. Luck of the draw meant that there are no Asian households among the Chosen 6(tm).

```{r}
maryland_ppp_naics_zcta %>% 
  filter(race=="Asian" & NHA ==0) %>% 
  group_by(zip5,NHA) %>% 
  summarize(
    total_loans=sum(amount),
    count_loans=n()
  ) %>% 
  arrange(desc(total_loans))

maryland_ppp_naics_zcta %>% 
  filter(race=="Asian" & zip5=="21031")

```

**Q7** How many industries have exactly one loan in the data (hint: you can filter _after_ using `summarise` on any totals you create). From a newsworthiness standpoint, which of these is most worthy of further exploration, and why?
**A7** There are 44 industries in Maryland that only had one business receiving a loan. Honestly, I think there are several industries here that could be newsworthy. If I had to pick one, I would want to know more about the company that builds missiles and spacecraft. The US Government certainly didn't slow down defense spending at any point during the pandemic, and I assume that's their only customer. Therefore, they shouldn't have seen a slowdown in business that mandated paycheck protection. I also know there was a huge ammunition shortage around the run-up to the 2020 Election (totally normal democracy), so I would be interested in how the ammo manufacturers were in such dire straits as to require large federal loans. 

```{r}
maryland_ppp_naics_zcta %>% 
  group_by(title) %>% 
  summarize(count_loans=n()) %>% 
  filter(count_loans==1)

maryland_ppp_naics_zcta %>% 
  filter(title=="Small Arms Ammunition Manufacturing")
```